### 2010-10-24 Thesis/Chapter6.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\begin{figure}[ht]
\includegraphics[width=13.9cm]{chapter6/sofa.pdf}
\caption [Multi-model representation in SOFA] {Multi-model representation in SOFA. \textbf{Left:} a Behaviour Model controls the other representations via a series of mappings. \textbf{Right:} examples of representations with a liver model.}
\label{chap6:fig-SOFA}
\end{figure}

@@

\subsubsection*{SOFA integration.}
Implementing a biomechanical model in SOFA translates essentially into writing a new Force Field, that is describing the algorithm used to compute internal forces in the model. It merely comes down to creating a single C++ class and changing the position reads and force writes to integrate the algorithm into SOFA's design. The precomputation phase takes place in the initialisation method where relevant variables are computed and passed to an external C function that allocates memory on the GPU and binds textures to it. During the simulation loop, the Solver requests the computation of the forces by launching the appropriate kernels on the GPU.

@@

In the precomputation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks).  Yet,  storing the force contributions in shared memory instead of global memory to read them later on via texture fetches  may  outperform the need of computing several elements several times.  However, the efficiency of this single kernel approach depends on the nature of the mesh. Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements). This type of mesh is often encountered when dealing with the complex shapes of anatomical structures. Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective.

@@

	\subsection{Pure shear of a cube}
The first example involved shearing of a unit cube model. We imposed a ramped loading such that the displacements $u$ of the loaded nodes at time $t$ were given by $u = rt$, where $r$ is the loading speed, and we chose the $x$-direction to be the preferred material direction ($\mathbf{a}_0 = [1\, 0\, 0 ]$). Since the deformation is homogenous only a single element was required to reproduce the stresses exactly. Moreover an analytical solution is available with which numerical results may be compared \citep{Taylor2009}. Four independent nonzero stress terms are produced.

@@

The end face reaction force history for each model is shown in \fig{chap6:fig-stressDecay}. A distinct decay curve resulting from stress relaxation, and similar to that commonly noted in biological tissues \citep{Fung93}, was observed. The increased stiffness afforded by the anisotropic models (albeit in an orthogonal direction to the loading) result in greater reaction forces than the isotropic model.
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=8cm]{chapter6/stressDecay.pdf}
\end{center}
\caption[FE computed reaction forces on the end faces of the cube models over time.]{FE computed reaction forces on the end faces of the cube models over time. Curves are labelled according to the constitutive model used, where \emph{Anisotropic 1} refers to the transversely isotropic viscoelastic model with $\mathbf{a}_0 = \mathbf{a}_0^y$ and \emph{Anisotropic 2} refers to the case of $\mathbf{a}_0 = \mathbf{a}_0^{yz}$. Note that in this figure both markers and solid lines correspond to FE solutions.}
\label{chap6:fig-stressDecay}
\end{figure}

@@

		\subsubsection*{General GPU performance}
Next, we assessed the computational performance of the GPU implementation and efficiency of the constitutive update scheme by measuring computation times for a range of mesh densities. Using the cube geometry as above we generated models with mesh sizes ranging from $3\,993$ DOF ($1\,331$ nodes, $1\,000$ hexahedral elements) to $177\,957$ DOF ($59\,319$ nodes, $54\,872$ hexahedral elements). We compared the GPU solution times for a single time step with those of a CPU-based implementation developed using C++. The test machine included an Intel Core2Duo 2.4GHz CPU, 2GB RAM, and an NVIDIA GeForce 8800GTX GPU. Three constitutive models were considered: the transversely isotropic viscoelastic model (TIV), TIV minus the viscoelastic terms (TIE), and TIE minus the anisotropy terms (NHE). Comparison of times for TIV and TIE indicates the computational load introduced by the developed viscoelastic constitutive update scheme. Comparison of times for TIE and NHE indicates the computational load introduced by material anisotropy. Figure~\ref{chap6:fig-timingsGPU} shows the results of the experiments. 	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU1.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU2.pdf}} \\
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU3.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU4.pdf}} \\
\caption[GPU solution times]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE. (b) also shows the critical time steps $ \Delta t_{cr} $ for each model size using liver-like and brain-like properties.} 
\label{chap6:fig-timingsGPU}
\end{figure}

In \fig{chap6:fig-timingsGPU}(a) and \fig{chap6:fig-timingsGPU}(b) it can be seen that GPU solution affords significant speed improvements over CPU solution for all constitutive models. Additionally it appears that solution times are little affected by the introduction of the more complex constitutive models, and importantly by the use of the developed constitutive update scheme. This is borne out in \fig{chap6:fig-timingsGPU}(d), where we observe that the maximum solution time ratio for model TIE to model NHE (anisotropy vs isotropy) was $1.013$, and that of model TIV to model TIE (viscoelastic vs elastic) was $1.043$. The largest total solution time increase for an anisotropic viscoelastic model compared with an isotropic hyperelastic one was $5.1\, \%$ We conclude that the key features of anisotropy and viscoelasticity may be included in simulations at very little additional computational cost. Referring to \fig{chap6:fig-timingsGPU}(c) we observe a maximum speed improvement for GPU solution over CPU solution of $56.3 \times$.

@@

The Von Mises equivalent stress distributions resulting from stretching the model with and without inclusion of viscoelasticity are compared in \fig{chap6:fig-liver}. For the viscoelastic case (\fig{chap6:fig-liver}, right image) the model is depicted several seconds after the completion of the stretch, so that stresses had reached an approximately steady state. The stress dissipation introduced by the viscoelastic model is clearly evident, and for example would result in significantly altered virtual tool reaction forces.	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver1.pdf}}
\hfill
\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver2.pdf}}
\hfill
\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver3.pdf}}
\caption[GPU solution times]{Deformation of a liver model using the TI model (b) and the TIV model (c). The undeformed mesh is shown in (a). Colour maps indicate the relative Von Mises equivalent stress magnitude.} 
\label{chap6:fig-liver}
\end{figure}

@@

Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, my efficient re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. My contribution was once again to implement these enhancements on GPU. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA. Over the course of my PhD, this work has led to three main publications \citep{Comas2008,Taylor2008,Taylor2009}.

>>>>

\begin{figure}[ht]
\includegraphics[width=13.9cm]{chapter6/sofa.pdf}
\caption [Multi-model representation in SOFA] {Multi-model representation in SOFA. The different representations are connected through a series of mappings (in red). Examples of representations with a liver model are also shown on the right.}
\label{chap6:fig-SOFA}
\end{figure}

@@

\subsubsection*{SOFA integration.}
Implementing a biomechanical model in SOFA translates essentially into writing a new Force Field, that is describing the algorithm used to compute internal forces in the model. It merely comes down to creating a single C++ class and changing the position reads and force writes to integrate the algorithm into SOFA's design. The pre-computation phase takes place in the initialisation method where relevant variables are computed and passed to an external C function that allocates memory on the GPU and binds textures to it. During the simulation loop, the Solver requests the computation of the forces by launching the appropriate kernels on the GPU.

@@

In the pre-computation phase, nodes and elements are sorted by block to leave enough space for storage in shared memory of the element force contributions initially calculated at the end of the first kernel. We add nodes to a block along with their unique elements until the shared memory is saturated. Of course, in this approach some elements will be computed several times (because being into different blocks).  Yet,  storing the force contributions in shared memory instead of global memory to read them later on via texture fetches  may  outperform the need of computing several elements several times.  However, the efficiency of this single kernel approach depends on the nature of the mesh. Because the force contributions of all elements attached to a node must be eventually added up, this approach allows to avoid the numerous (slow) accesses to global memory when the valency of the nodes is high (that is, nodes are attached to many elements). This type of mesh is often encountered when dealing with the complex shapes of anatomical structures. Therefore, where the first implementation showed its limits with complex meshes, this single kernel approach has proven to be more computationally effective.

@@

	\subsection{Pure shear of a cube}
The first example involved shearing of a unit cube model. We imposed a ramped loading such that the displacements $u$ of the loaded nodes at time $t$ were given by $u = rt$, where $r$ is the loading speed, and we chose the $x$-direction to be the preferred material direction ($\mathbf{a}_0 = [1\, 0\, 0 ]$). Since the deformation is homogeneous only a single element was required to reproduce the stresses exactly. Moreover an analytical solution is available with which numerical results may be compared \citep{Taylor2009}. Four independent nonzero stress terms are produced.

@@

The end face reaction force history for each model is shown in \fig{chap6:fig-stressDecay}. A distinct decay curve resulting from stress relaxation, and similar to that commonly noted in biological tissues \citep{Fung93}, was observed. The increased stiffness afforded by the anisotropic models (albeit in an orthogonal direction to the loading) result in greater reaction forces than the isotropic model.
%
\begin{figure}[ht]
\begin{center}
\includegraphics[width=8cm]{chapter6/stressDecay.pdf}
\end{center}
\caption[Finite element computed reaction forces on the end faces of the cube models over time.]{Finite element computed reaction forces on the end faces of the cube models over time. Curves are labelled according to the constitutive model used, where \emph{Anisotropic 1} refers to the transversely isotropic viscoelastic model with $\mathbf{a}_0 = \mathbf{a}_0^y$ and \emph{Anisotropic 2} refers to the case of $\mathbf{a}_0 = \mathbf{a}_0^{yz}$. Note that in this figure both markers and solid lines correspond to finite element solutions.}
\label{chap6:fig-stressDecay}
\end{figure}

@@

		\subsubsection*{General GPU performance}
Next, we assessed the computational performance of the GPU implementation and efficiency of the constitutive update scheme by measuring computation times for a range of mesh densities. Using the cube geometry as above we generated models with mesh sizes ranging from $3\,993$ DOF ($1\,331$ nodes, $1\,000$ hexahedral elements) to $177\,957$ DOF ($59\,319$ nodes, $54\,872$ hexahedral elements). We compared the GPU solution times for a single time step with those of a CPU-based implementation developed using C++. The test machine included an Intel Core2Duo 2.4GHz CPU, 2GB RAM, and an NVIDIA GeForce 8800GTX GPU. Three constitutive models were considered: the transversely isotropic viscoelastic model (TIV), TIV minus the viscoelastic terms (TIE), and TIE minus the anisotropy terms (NHE). Comparison of times for TIV and TIE indicates the computational load introduced by the developed viscoelastic constitutive update scheme. Comparison of times for TIE and NHE indicates the computational load introduced by material anisotropy. Figure~\ref{chap6:fig-timingsGPU} shows the results of the experiments. 	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU1.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU2.pdf}} \\
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU3.pdf}}
\hspace{0.5cm}
\subfloat[ ]{\includegraphics[width=6.5cm]{chapter6/timingsGPU4.pdf}} \\
\caption[GPU solution times for different constitutive models]{(a), (b) show CPU and GPU solution times, respectively, for a single time step for each constitutive model as a function of model size, (c) shows the ratio of CPU to GPU solution times for model TIV, and (d) shows ratios of GPU solution times for the constitutive models: TIV to TIE, TIE to NHE, and TIV to NHE. (b) also shows the critical time steps $ \Delta t_{cr} $ for each model size using liver-like and brain-like properties.} 
\label{chap6:fig-timingsGPU}
\end{figure}

In \fig{chap6:fig-timingsGPU}(a) and \fig{chap6:fig-timingsGPU}(b) it can be seen that GPU solution affords significant speed improvements over CPU solution for all constitutive models. Additionally it appears that solution times are little affected by the introduction of the more complex constitutive models, and importantly by the use of the developed constitutive update scheme. This is borne out in \fig{chap6:fig-timingsGPU}(d), where we observe that the maximum solution time ratio for model TIE to model NHE (anisotropy versus isotropy) was $1.013$, and that of model TIV to model TIE (viscoelastic versus elastic) was $1.043$. The largest total solution time increase for an anisotropic viscoelastic model compared with an isotropic hyperelastic one was $5.1\, \%$ We conclude that the key features of anisotropy and viscoelasticity may be included in simulations at very little additional computational cost. Referring to \fig{chap6:fig-timingsGPU}(c) we observe a maximum speed improvement for GPU solution over CPU solution of $56.3 \times$.

@@

The Von Mises equivalent stress distributions resulting from stretching the model with and without inclusion of viscoelasticity are compared in \fig{chap6:fig-liver}. For the viscoelastic case (\fig{chap6:fig-liver}, right image) the model is depicted several seconds after the completion of the stretch, so that stresses had reached an approximately steady state. The stress dissipation introduced by the viscoelastic model is clearly evident, and for example would result in significantly altered virtual tool reaction forces.	
%
\begin{figure}[ht]
\centering 
\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver1.pdf}}
\hfill
\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver2.pdf}}
\hfill
\subfloat[ ]{\includegraphics[width=4cm]{chapter6/liver3.pdf}}
\caption[Deformation of a liver using TI and TIV models]{Deformation of a liver model using the TI model (b) and the TIV model (c). The undeformed mesh is shown in (a). Colour maps indicate the relative Von Mises equivalent stress magnitude.} 
\label{chap6:fig-liver}
\end{figure}

@@

Hence, provided that the application is appropriate, the TLED algorithm is an efficient and accurate algorithm to model the deformation of solid soft anatomical organs. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, my efficient re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up to $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. My contribution was once again to implement these enhancements on GPU. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publicly with the upcoming 1.0 release of SOFA. Over the course of my PhD, this work has led to three main publications \citep{Comas2008,Taylor2008,Taylor2009}.

### 2010-10-24 Thesis/Chapter7.tex
- Update 24 October (final) -
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@189 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter[Modelling the deformation of hollow objects in real-time][Modelling hollow objects in real-time]{Modelling the deformation of hollow objects in real-time}
\label{chap7}
\begin{shortAbstract}
The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging. In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs (like brain, liver, prostate etc.). But there is also a need to simulate the deformation of hollow organs, whose volume is negligible compared to their surface area, such as colon, blood vessels, stomach etc. The solid models previously mentioned are not appropriate and specific approaches are required for these thin anatomical structures. In this chapter, we will review diverse techniques applied for the simulation of thin objects. These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a non-physical bending energy and (3) methods based on the equations of continuum mechanics. Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible. 
\end{shortAbstract}

@@

An obvious modelling for thin structures is to apply the same methods than with solid objects. For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an arterial wall for simulating an angioplasty procedure. Their model is non-linear and anisotropic along fibres placed according to histology. They also model diseased part of the vessel through a hardening parameter which adds plasticity. However, they discretised their geometric model with eight-node brick elements. Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object. If this approach can work, about a dozen of elements is required within the thickness of the structure for the system to be numerically stable \TODO{reference?}. Consequently, many elements are needed to model thin structures and real-time performance is difficult to achieve. Hence, specific models for thin objects were developed.

@@

Early approaches to thin objects modelling only considered in-plane deformation, and often relied on mass-spring models. This technique was already presented in section \ref{chap4:massSprings} so we only present their use for modelling thin structures. \cite{Provot95} improved a mass-spring model to take into account the non-elastic properties of woven fabrics. Because of the high stiffness of textiles, mass-spring models are particularly unstable when used to simulate cloth. The author proposed a new method inspired from dynamic inverse procedure. Under certain conditions and because of the linear law employed in a mass-spring system, the fabrics may become super-elastic. \citeauthor{Provot95} suggested to set a threshold on the deformation rate to correct this super-elasticity (see \fig{chap7:fig-sheet}). 
%
\begin{figure}[ht]
\centering 
\subfloat[Initial position]{\includegraphics[width=4.5cm]{chapter7/sheet1.pdf}}
\hfill 
\subfloat[Uncorrected]{\includegraphics[width=4.5cm]{chapter7/sheet2.pdf}}
\hfill 
\subfloat[Corrected]{\includegraphics[width=4.5cm]{chapter7/sheet3.pdf}}
\caption{Deformation of a sheet hanging by two adjacent corners. Images courtesy of \cite{Provot95}.}
\label{chap7:fig-sheet}
\end{figure}

@@

\section{Objectives}

We seek to propose a solution for simulating, in real-time, the deformation of thin anatomical structures. Since \cite{Black91} showed that formulations with only membrane energy were likely to produce significant errors in the stress distribution, this model is required to feature bending. Ideally, this solution would be versatile and allow us to simulate thin objects with various shapes and material properties with good accuracy. In this regard, methods derived from continuum mechanics are the most appropriate. In some situations, a thin plate model would not correctly capture the deformation of some objects and therefore a shell formulation was retained for more flexibility. To the best of our knowledge, no real-time shell finite element method was presented in the field of medical simulation.

In the next section, we will first introduce our co-rotational triangular shell finite element model and show how to efficiently apply this model to simulate the implant deployment in cataract surgery. In a second time, we will present a technique to mesh complex anatomical shapes with curved triangles, hence allowing us to model, in real-time, any anatomical structures with our co-rotational shell FEM in an optimal way.

>>>>

\chapter{Modelling the deformation of hollow objects in real-time}
\label{chap7}
\begin{shortAbstract}
The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging. In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs (like brain, liver, prostate etc.). But there is also a need to simulate the deformation of hollow organs, whose volume is negligible compared to their surface area, such as colon, blood vessels, stomach etc. The solid models previously mentioned are not appropriate and specific approaches are required for these thin anatomical structures. In this chapter, we will review diverse techniques applied for the simulation of thin objects. These different approaches may be grouped into three main categories: (1) mass-spring models, (2) approaches which derive a non-physical bending energy and (3) methods based on the equations of continuum mechanics. Although very few models have been proposed for simulating the deformation of thin anatomical structures in the field of medical simulation, we will try to emphasize medical applications whenever possible. 
\end{shortAbstract}

@@

An obvious modelling for thin structures is to apply the same methods than with solid objects. For instance, \cite{Holzapfel02} proposed a very complex 3-layer model of an arterial wall for simulating an angioplasty procedure. Their model is non-linear and anisotropic along fibres placed according to histology. They also model diseased part of the vessel through a hardening parameter which adds plasticity. However, they discretised their geometric model with eight-node brick elements. Another example is given by \cite{Aloisio04} who assume that the blood flow in the vessels may be considered as incompressible and the artery consequently becomes a solid body and can be modelled as a solid deformable object. If this approach can work, about a dozen of elements is required within the thickness of the structure for the system to be numerically stable. Consequently, many elements are needed to model thin structures and real-time performance is difficult to achieve. Hence, specific models for thin objects were developed.

@@

Early approaches to thin objects modelling only considered in-plane deformation, and often relied on mass-spring models. This technique was already presented in section \ref{chap4:massSprings} so we only present their use for modelling thin structures. \cite{Provot95} improved a mass-spring model to take into account the non-elastic properties of woven fabrics. Because of the high stiffness of textiles, mass-spring models are particularly unstable when used to simulate cloth. The author proposed a new method inspired from dynamic inverse procedure. Under certain conditions and because of the linear law employed in a mass-spring system, the fabrics may become super-elastic. \citeauthor{Provot95} suggested to set a threshold on the deformation rate to correct this super-elasticity (see \fig{chap7:fig-sheet}). 
%
\begin{figure}[ht]
\centering 
\subfloat[Initial position]{\includegraphics[width=4.5cm]{chapter7/sheet1.pdf}}
\hfill 
\subfloat[Uncorrected]{\includegraphics[width=4.5cm]{chapter7/sheet2.pdf}}
\hfill 
\subfloat[Corrected]{\includegraphics[width=4.5cm]{chapter7/sheet3.pdf}}
\caption[Deformation of a sheet hanging by two adjacent corners]{Deformation of a sheet hanging by two adjacent corners. Images courtesy of \cite{Provot95}.}
\label{chap7:fig-sheet}
\end{figure}

@@

\section{Conclusion}

The human body is composed of various deformable anatomical structures and finding appropriate soft-tissue models for all these type of structures is challenging. In the previous two chapters, we described an efficient non-linear FEM formulation for simulating the deformation of solid organs. However, the solid models previously mentioned are not appropriate and specific approaches are required for thin anatomical structures. In this chapter, we reviewed diverse techniques applied for the simulation of thin objects. It appears that few techniques derived from continuum mechanics have been proposed for modelling in real-time hollow structures in medical simulation.

Therefore, we seek to propose a solution for simulating the deformation of thin anatomical structures in a physically realistic manner. In addition, this approach will have to be computable in real-time. Since \cite{Black91} showed that formulations with only membrane energy were likely to produce significant errors in the stress distribution, this model is required to feature bending. Ideally, this solution would be versatile and allow us to simulate thin objects with various shapes and material properties with good accuracy. In this regard, methods derived from continuum mechanics are the most appropriate. In some situations, a thin plate model would not correctly capture the deformation of some objects and therefore a shell formulation was retained for more flexibility. To the best of our knowledge, no real-time shell finite element method was presented in the field of medical simulation.

In the next chapter, we will first introduce our co-rotational triangular shell finite element model and show how to efficiently apply this model to simulate the implant deployment in cataract surgery (see chapter~\ref{chap8}). In a second time, we will present a technique to mesh complex anatomical shapes with curved triangles, hence allowing us to model, in real-time, any anatomical structures with our co-rotational shell FEM in an optimal way (see chapter~\ref{chap9}).
